---
title: "Practicum I CS5200"
author: "Gowreesh Gunupati, Vasumathi Narayanan"
date: "Spring 2023"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

## Question 3 - Connect to Database
We created a new database in MySQL using command "CREATE DATABASE birdstrikes" and we connected the MySQL server to R using workbench.
We are using library lubridate for parsing and manipulating the date. 

```{r connectToDatabase, eval=TRUE, warning=FALSE}
library(RMySQL)
library(sqldf)
library(lubridate)
birdStrikesDBCon = dbConnect(RMySQL::MySQL(),
                            dbname='birdstrikes',
                            host='localhost',
                            port=3306,
                            user='ecommerceapp',
                            password='ecommerceapp')
```

setting local_infile to True

```{sql settinglocalinfile,connection=birdStrikesDBCon}
#show global variables like 'local_infile';
SET global local_infile = TRUE;
```

This is loaded to use sqldf for loading the data frame with respective columns using SQLite.

```{r loadingsqldriver, eval=TRUE, warning=FALSE}
options(sqldf.driver = "SQLite")
```

## Drop If Tables already Exist

```{sql Droptableifexistsincidents,connection=birdStrikesDBCon}
DROP TABLE IF EXISTS incidents;
```
```{sql Droptableairports,connection=birdStrikesDBCon}
DROP TABLE IF EXISTS airports;
```
```{sql Droptableifexistsconditions,connection=birdStrikesDBCon}
DROP TABLE IF EXISTS conditions;
```
```{sql Droptableifexistsairlines,connection=birdStrikesDBCon}
DROP TABLE IF EXISTS airlines;
```



We have created the airports table with the appropriate attributes and the constraints are mentioned as below. aid is a synthetic primary key, so we have added AUTO_INCREMENT for that.

```{sql createTableAirportForDB, connection = birdStrikesDBCon}
CREATE TABLE airports (
  aid INT AUTO_INCREMENT,
  airportName VARCHAR(100) ,
  airportCode VARCHAR(10),
  state VARCHAR(100),
  PRIMARY KEY(aid)
  ); 
```

We have created the conditions table with the appropriate attributes and the constraints are mentioned as below. cid is the primary key here.

```{sql createTableConditionsForDB, connection = birdStrikesDBCon}
CREATE TABLE conditions (
cid INT(10) AUTO_INCREMENT,
skyCondition VARCHAR(50),
explaination VARCHAR(50),
PRIMARY KEY(cid)
);
```

We have created the airlines table with the appropriate attributes and the constraints are mentioned as below. "eid" is the synthetic primary key so we have added AUTO_INCREMENT.

```{sql createTableAirlinesForDB, connection = birdStrikesDBCon}
CREATE TABLE airlines (
eid INT AUTO_INCREMENT, 
airlineName VARCHAR(50),
airlineCode VARCHAR(50),
flag VARCHAR(10),
PRIMARY KEY(eid)
); 
```

We have created the incidents table with the appropriate attributes and the constraints are mentioned as below. 
origin is the foreign key that is referenced to aid
conditions is the foreign key that is referenced to conditions
airline is the foreign key that is reference to airlines

```{sql createTableIncidentsForDB, connection = birdStrikesDBCon}
CREATE TABLE incidents(
  rid INT,
  depDate DATE,
  origin INT NOT NULL ,
  airline INT NOT NULL,
  aircraft VARCHAR(50),
  flightPhase VARCHAR(20),
  altitude INT,
  conditions INT NOT NULL,
  warned TINYINT,
  PRIMARY KEY (rid),
  CONSTRAINT CHK_altitude CHECK (altitude>=0),
  FOREIGN KEY (origin) REFERENCES airports(aid),
  FOREIGN KEY (conditions) REFERENCES conditions(cid),
  FOREIGN KEY (airline) REFERENCES airlines(eid),
  CONSTRAINT CHK_flightPhase CHECK 
  (flightPhase = 'takeoff' OR 
  flightPhase='landing' OR 
  flightPhase='inflight' OR 
  flightPhase ='unknown')
  );
  
  
```
## Verifying Schemas for birdstrikes

```{sql checkIncientsSchemas, connection = birdStrikesDBCon}
DESC  incidents;
```

```{sql checkAirportsSchemas, connection = birdStrikesDBCon}
DESC airports;
```
```{sql checkConditionsSchemas, connection = birdStrikesDBCon}

DESC conditions;
```
```{sql checkAirlinesSchemas, connection = birdStrikesDBCon}
DESC airlines;
```


## Question 5 Loading CSV to R Dataframe
```{r loadingBirdStrikesToDataframe, eval=TRUE, warning=FALSE}
bds.raw<- read.csv("BirdStrikesData-V2.csv", header = TRUE, sep = ",")
head(bds.raw)
```

## Data Cleaning and pre-processing to format the date

1. We are normalizing the flight phases to have only values takeoff, landing, inflight,unknown
2. The flight_date field was transformed into %Y-%m-%d
3. Changed all the flights, airline, aircrafts that holds the value as empty string to unknown

## Parsing Date
```{r parseDateforDeparture, eval=TRUE, warning=FALSE}
bds.raw$flight_date<- parse_date_time(bds.raw$flight_date, orders=c("m-d-y H:M","m/d/y H:M"))
bds.raw$flight_date <- format(bds.raw$flight_date,"%Y-%m-%d")
```

## Data cleaning to rename empty values to unknown
```{r cleaningSentinelData, eval=TRUE, warning=FALSE}
case5 <- which(bds.raw$airline==''| bds.raw$airline==' '| 
                 bds.raw$airline=='NA' | bds.raw$airline=='N/A')
bds.raw[case5,"airline"] <- 'unknown'

case6 <- which(bds.raw$aircraft==''|bds.raw$aircraft==' '| 
                 bds.raw$aircraft=='NA'| bds.raw$aircraft=='N/A' )
bds.raw[case5,"aircraft"] <- 'unknown'
```

## Harmonizing the flight phases to be one of takeoff, landing, inflight, unknown.
```{r harmonizingFlightPhases, eval=TRUE, warning=FALSE}
case1 <- which(bds.raw$flight_phase=='Take-off run' 
               | bds.raw$flight_phase=='Climb' )
bds.raw[case1,"flight_phase"] <- 'takeoff'

case2 <- which(bds.raw$flight_phase=='Landing Roll'
               |bds.raw$flight_phase=='Taxi'|bds.raw$flight_phase=='Parked')
bds.raw[case2,"flight_phase"] <- 'landing'

case3 <-which(bds.raw$flight_phase=='Approach' | bds.raw$flight_phase=='Descent')
bds.raw[case3,"flight_phase"] = 'inflight'

case4 <- which(bds.raw$flight_phase=='' | bds.raw$flight_phase=='N/A' 
               | bds.raw$flight_phase=='NA' | bds.raw$flight_phase==' ')
bds.raw[case4,"flight_phase"] <- 'unknown'
```

## Change the warned flag as tinyint datatype based on pilot_warned_flag
```{r changingFlags, eval=TRUE, warning=FALSE}
warned <- which(bds.raw$pilot_warned_flag=="Y")
notwarned <- which(bds.raw$pilot_warned_flag=='N')

bds.raw[warned,"pilot_warned_flag"] <- 1
bds.raw[notwarned,"pilot_warned_flag"] <- 0
```


## setting default sentinel values for model, origin and airport
```{r changetoUnknown, eval=TRUE,warning=FALSE}
case8 <- which(bds.raw$airport==''| bds.raw$airport==' '  
               | bds.raw$airport=='NA' |bds.raw$airport=='N/A')
bds.raw[case8,"airport"]<- 'unknown'
case9 <- which(bds.raw$model=='' | bds.raw$model==' '
               |bds.raw$model=='NA' |bds.raw$model=='N/A')
bds.raw[case9,"model"]<- 'unknown'
case10<- which(bds.raw$origin=='NA'| bds.raw$origin==''
               |bds.raw$origin==' '| bds.raw$origin=='N/A')
bds.raw[case10,"origin"]<-'unknown'
```

## omit flights without flight information
```{r omitEmptyFlightDetails, eval=TRUE, warning=FALSE}
bds.raw <- bds.raw[which(bds.raw$airport!='' & bds.raw$model!='' & bds.raw$origin!='' ),]
```

## Question 6 Loading R Dataframe to bird-strikes-DB

### Inserting into conditions table

```{r insertToairportsTable, eval=TRUE,warning=FALSE}
conditionsdata <- sqldf::sqldf("SELECT distinct(sky_conditions)
                               AS skyCondition FROM [bds.raw]");
dbWriteTable(birdStrikesDBCon, "conditions", conditionsdata, 
             append = TRUE, row.names = FALSE)
n.conditionsdata <- nrow(conditionsdata)
conditionsdata$cid <- seq(1, n.conditionsdata)
```
```{r conditionsdata,eval = F}
head(conditionsdata)
```

### Inserting into airlines table
```{r insertToairlineTable, eval=TRUE, warning=FALSE}
airlinesData <- sqldf::sqldf("SELECT DISTINCT(TRIM(airline)) AS airlineName,
                             pilot_warned_flag AS flag  from [bds.raw]");
dbWriteTable(birdStrikesDBCon, "airlines", airlinesData, append = TRUE, row.names = FALSE)
n.airlinesData <- nrow(airlinesData)
airlinesData$eid <- seq(1, n.airlinesData)
```

```{r airlinesData,eval = F}
head(airlinesData)
```

### Inserting into airport table

```{r insertToairportTable, eval=TRUE,warning=FALSE}
airportsdata <- sqldf::sqldf("SELECT DISTINCT(trim(airport)) AS airportName, 
                             origin AS state from [bds.raw]")
dbWriteTable(birdStrikesDBCon, "airports", airportsdata, append = TRUE, row.names = FALSE)
n.airportsdata <- nrow(airportsdata)
airportsdata$aid <- seq(1, n.airportsdata)
```
```{r airportsdata,eval = F}
head(airportsdata)
```
### Inserting into incidents table
```{r insertToIncidentsTable, eval=TRUE,warning=FALSE}
incidentsdata <- sqldf::sqldf("SELECT rid, flight_date as depDate, origin,
                              airport, airline, aircraft,flight_phase as flightPhase,
                              altitude_ft as altitude, sky_conditions as conditions, 
                              pilot_warned_flag as warned FROM [bds.raw]");
n.incidentsdata <- nrow(incidentsdata)
for(r in 1:n.incidentsdata)
{
  
  a <- airportsdata$aid[which(airportsdata$airportName == incidentsdata$airport[r])]
  incidentsdata$origin[r] <- a[[1]][1]
  
  b <- conditionsdata$cid[which(conditionsdata$skyCondition == incidentsdata$conditions[r])]
  incidentsdata$conditions[r] <- b[[1]][1]
  
  c<-airlinesData$eid[which(airlinesData$airlineName == incidentsdata$airline[r])] 
  incidentsdata$airline[r] <- c[[1]][1]
  
}

incidentsdata <- sqldf::sqldf("SELECT rid, depDate , origin, airline,
                              aircraft,flightPhase, altitude, conditions,
                              warned FROM incidentsdata");

dbWriteTable(birdStrikesDBCon, "incidents", incidentsdata, append = TRUE, row.names = FALSE)

```
```{r incidentsdata,eval = F}
head(incidentsdata)
```

1.Since birdstrikes dataframe is too large, we are splitting it into airportsdata, airlinesData,
incidentsdata and conditionsdata in order to make the data easier to handle.
2.We have then proceeded to populate the dataframes into the respective sql tables.



## Question 7 Verifying the table populations

```{sql conditionsselection, connection=birdStrikesDBCon}
SELECT * from conditions limit 5;
```
```{sql airportsselection, connection=birdStrikesDBCon}
SELECT * from airports limit 5;
```
```{sql airlineselection, connection=birdStrikesDBCon}
SELECT * from airlines limit 5;
```
```{sql incidentsselection, connection=birdStrikesDBCon}
SELECT * from incidents limit 5;
```
Test query to show all the foreign key constraints work properly after data loaded into their respective Tables
## Test query
```{sql testQuery1, connection=birdStrikesDBCon}
SELECT incidents.rid,incidents.depDate, airports.airportName AS origin,
airlines.airlineName AS airline, incidents.aircraft, 
incidents.flightPhase, incidents.altitude, 
conditions.skyCondition AS conditions, incidents.warned 
FROM incidents,airports, airlines,conditions 
WHERE rid = 1195
AND airports.aid = incidents.origin 
AND airlines.eid = incidents.airline
AND conditions.cid= incidents.conditions
```

This test query is to show that airline has UNKNOWN value in its column whose incident rid is 263834
```{sql testQuery2, connection=birdStrikesDBCon}
SELECT incidents.rid,incidents.depDate, airports.airportName AS origin,
airlines.airlineName AS airline, incidents.aircraft, 
incidents.flightPhase, incidents.altitude, 
conditions.skyCondition AS conditions, incidents.warned 
FROM incidents,airports, airlines,conditions 
WHERE rid=263834
AND airports.aid = incidents.origin 
AND airlines.eid = incidents.airline
AND conditions.cid= incidents.conditions
```
## Question 8
```{sql stateandNumberofincidents,connection=birdStrikesDBCon}
SELECT airports.state, COUNT(incidents.rid) AS "Number of Incidents" 
FROM incidents,airports  
WHERE airports.aid= incidents.origin 
GROUP BY (airports.state) 
ORDER BY COUNT(incidents.rid) 
DESC LIMIT 10;
```

## Question 9
```{sql question9,connection=birdStrikesDBCon}
SELECT DISTINCT  airlines.airlineName, COUNT(incidents.rid) AS "num_incidents"
FROM incidents,airlines 
WHERE airlines.eid= incidents.airline
GROUP BY (airlines.airlineName) HAVING num_incidents>
(SELECT AVG(num_incidents) FROM (SELECT DISTINCT  airlines.airlineName, 
COUNT(incidents.rid) AS "num_incidents"
FROM incidents,airlines 
WHERE airlines.eid= incidents.airline
GROUP BY airlines.airlineName) AS average);
```

<!--
Group by months and flight phase separately.
-->
```{r groupByMonthAndFlightPhase, eval=TRUE, warning=FALSE}
no_of_incidents_group_by_month<- dbGetQuery(birdStrikesDBCon,"SELECT DISTINCT MONTH(depDate) AS months ,
                                            COUNT(rid) as num_of_incidents_by_month FROM incidents 
                                            GROUP BY MONTH(depDate) ORDER BY  MONTH(depDate)")
no_of_incidents_group_by_flightphase <-dbGetQuery(birdStrikesDBCon,"SELECT DISTINCT incidents.flightPhase , COUNT(rid) AS num_of_incidents_by_flightphase FROM incidents GROUP BY incidents.flightPhase")
```
```{r returninghead, eval=TRUE,warning=FALSE}
head(no_of_incidents_group_by_month,6)
head(no_of_incidents_group_by_flightphase,6)
```

## Question 10
<!--
Group by months and flight phase together
-->
```{r groupByMonth_FlightPhase, eval=TRUE, warning=FALSE}
no_of_incidents_group_by_month_flightPhase<- 
  dbGetQuery(birdStrikesDBCon,"select distinct flightPhase,MONTH(depDate) as months ,
             COUNT(rid) as num_of_incidents_by_month 
             from incidents group by MONTH(depDate),
             flightPhase ORDER BY  MONTH(depDate)")
no_of_incidents_group_by_month_flightPhase=no_of_incidents_group_by_month_flightPhase[which(no_of_incidents_group_by_month_flightPhase$months !='NA'),]
head(no_of_incidents_group_by_month_flightPhase,6)
```
## Question 11
```{r plotNoOfIncidentsVsMonths, fig.height = 4, eval=TRUE, warning=FALSE}
plot( 
  x = no_of_incidents_group_by_month_flightPhase$months,y = 
       no_of_incidents_group_by_month_flightPhase$num_of_incidents_by_month
,
   xlab = "Month",
   ylab = "No of Bird Strikes",
   main = "Flight phase wise - Month vs Number of Bird Strike Incidents",
   col =  c("blue","green","red"),
   pch=19)

text(no_of_incidents_group_by_month_flightPhase$months,
     no_of_incidents_group_by_month_flightPhase$num_of_incidents_by_month, labels=no_of_incidents_group_by_month_flightPhase$num_of_incidents_by_month,
     cex = 0.5, pos = 4)

legend(x = "topright", legend=unique(no_of_incidents_group_by_month_flightPhase$flightPhase), 
       col = c("blue","green","red"), pch=19 )


```

## Question 12

```{sql doppingtrigger,connection=birdStrikesDBCon}
drop procedure if exists add_incident;
```

```{sql creatingprocedure,connection=birdStrikesDBCon}
create procedure add_incident(
in airport1 text, 
in state1 text, 
in date1 date, 
in airline1 text, 
in aircraft1 text, 
in flightPhase1 text, 
in altitude1 int, 
in condition1 text, 
in warning1 tinyint )

begin

declare cid1 int;
declare aid1 int default 0;
declare rid1 int;
declare eid1 int;

SELECT cid INTO cid1 FROM conditions WHERE conditions.skyCondition = condition1 ;

SELECT aid INTO aid1 FROM airports WHERE airportName = airport1 AND state = state1 ;

SELECT eid INTO eid1 FROM airlines WHERE airlineName= airline1 and flag=warning1;

SELECT (MAX(rid) + 1) INTO rid1 FROM incidents ;

IF aid1 = 0 THEN
SELECT (MAX(aid) + 1) INTO aid1 FROM airports ;
  INSERT INTO airports(aid, airportName, state) VALUES ( aid1, airport1, state1 ) ;
  END IF;

INSERT INTO incidents(rid, depDate,origin, airline, aircraft, 
flightPhase, altitude, conditions, warned) 
VALUES 
( rid1, date1, aid1, eid1, aircraft1, flightPhase1, altitude1, cid1, warning1 ) ;
end
```



```{sql callProcedure, connection=birdStrikesDBCon}
CALL add_incident(
'Newyork Airport',
'New York',
'2022-01-01',
'BUSINESS',
'Airplane',
'takeoff',
363,
'Overcast',
0);
```


```{sql tesstcase, connection=birdStrikesDBCon}
SELECT *FROM incidents WHERE depDate = '2022-01-01' 
```
From the below query we can see that the foreign key is used to retrieve the actual origin name, airline name and sky conditions by mapping the id.
This is same as what we gave as the input for the stored procedure.

```{sql connection=birdStrikesDBCon}
SELECT incidents.rid,incidents.depDate, airports.airportName AS origin,
airlines.airlineName AS airline, incidents.aircraft, 
incidents.flightPhase, incidents.altitude, 
conditions.skyCondition AS conditions, incidents.warned 
FROM incidents,airports, airlines,conditions 
WHERE depDate = '2022-01-01' 
AND airports.aid = incidents.origin 
AND airlines.eid = incidents.airline
AND conditions.cid= incidents.conditions
```

```{sql checkingAirportName,connection=birdStrikesDBCon}
SELECT * FROM airports WHERE airportName = 'Newyork Airport'
```
## Disconnecting database
```{r disconnectFromDB,eval=TRUE, warning=FALSE}
dbDisconnect(birdStrikesDBCon)
```
