---
title: "Practicum I CS5200"
author: "Gowreesh Gunupati, Vasumathi Narayanan"
date: "Spring 2023"
output: pdf_document
---

## Question 3 - Connect to Database

```{r connectToDatabase}̥
library(RMySQL)
library(sqldf)
library(lubridate)
birdStrikesDBCon = dbConnect(RMySQL::MySQL(),
                            dbname='birdstrikes',
                            host='localhost',
                            port=3306,
                            user='ecommerceapp',
                            password='ecommerceapp')
```

```{sql connection=birdStrikesDBCon}
#show global variables like 'local_infile';
set global local_infile = TRUE;
```


This is loaded to use sqldf for loading the data frame with respective columns.

```{r}
options(sqldf.driver = "SQLite")
```

## Drop If Tables already Exist

```{sql connection=birdStrikesDBCon}
DROP TABLE IF EXISTS incidents;
```
```{sql connection=birdStrikesDBCon}
DROP TABLE IF EXISTS airports;
```
```{sql connection=birdStrikesDBCon}
DROP TABLE IF EXISTS conditions;
```
```{sql connection=birdStrikesDBCon}
DROP TABLE IF EXISTS airlines;
```



We have created the airports table with the appropriate attributes and the constraints are mentioned as below. aid is a synthetic primary key, so we have added AUTO_INCREMENT for that.

```{sql createTableAirportForDB, connection = birdStrikesDBCon}
CREATE TABLE airports (
  aid INT AUTO_INCREMENT,
  airportName VARCHAR(100) ,
  airportCode VARCHAR(10),
  state VARCHAR(100),
  PRIMARY KEY(aid)
  ); 
```

We have created the conditions table with the appropriate attributes and the constraints are mentioned as below. cid is the primary key here.

```{sql createTableConditionsForDB, connection = birdStrikesDBCon}
CREATE TABLE conditions (
cid INT(10) AUTO_INCREMENT,
skyCondition VARCHAR(50),
explaination VARCHAR(50),
PRIMARY KEY(cid)
);
```

We have created the airlines table with the appropriate attributes and the constraints are mentioned as below. "eid" is the synthetic primary key so we have added AUTO_INCREMENT.

```{sql createTableAirlinesForDB, connection = birdStrikesDBCon}
CREATE TABLE airlines (
eid INT AUTO_INCREMENT, 
airlineName VARCHAR(50),
airlineCode VARCHAR(50),
flag VARCHAR(10),
PRIMARY KEY(eid)
); 
```

We have created the incidents table with the appropriate attributes and the constraints are mentioned as below. 

```{sql createTableIncidentsForDB, connection = birdStrikesDBCon}
CREATE TABLE incidents(
  rid INT,
  depDate DATE,
  origin INT NOT NULL ,
  airline INT NOT NULL,
  aircraft VARCHAR(50),
  flightPhase VARCHAR(20),
  altitude INT,
  conditions INT NOT NULL,
  warned TINYINT,
  PRIMARY KEY (rid),
  CONSTRAINT CHK_altitude CHECK (altitude>=0),
  FOREIGN KEY (origin) REFERENCES airports(aid),
  FOREIGN KEY (conditions) REFERENCES conditions(cid),
  FOREIGN KEY (airline) REFERENCES airlines(eid),
  CONSTRAINT CHK_flightPhase CHECK (flightPhase = 'takeoff' OR flightPhase='landing' OR flightPhase='inflight' OR flightPhase ='unknown')
  );
  
  
```
## Verifying Schemas for birdstrikes

```{sql checkIncientsSchemas, connection = birdStrikesDBCon, eval =F}
DESC  incidents;
```

```{sql checkAirportsSchemas, connection = birdStrikesDBCon, eval =F}
DESC airports;
```
```{sql checkConditionsSchemas, connection = birdStrikesDBCon, eval =F}

DESC conditions;
```
```{sql checkAirlinesSchemas, connection = birdStrikesDBCon, eval =F}
DESC airlines;
```


## Question 5 Loading CSV to R Dataframe
```{r loadingBirdStrikesToDataframe}
bds.raw<- read.csv("BirdStrikesData-V2.csv", header = TRUE, sep = ",")
head(bds.raw)
```

## Data Cleaning and pre-processing to format the date

1. We are normalizing the flight phases to have only values takeoff,landing,inflight,unknown
2. The flight_date field was transformed into
```{r parseDateforDeparture}
bds.raw$flight_date<- parse_date_time(bds.raw$flight_date, orders=c("m-d-y H:M","m/d/y H:M"))
bds.raw$flight_date <- format(bds.raw$flight_date,"%Y-%m-%d")
```
```{r cleaningSentinelData}
case5 <- which(bds.raw$airline=='')
bds.raw[case5,"airline"] <- 'unknown'

case6 <- which(bds.raw$aircraft=='')
bds.raw[case5,"aircraft"] <- 'unknown'
```
## Harmonizing the flight phases to be one of takeoff, landing, inflight, unknown.
```{r harmonizingFlightPhases}
case1 <- which(bds.raw$flight_phase=='Take-off run' | bds.raw$flight_phase=='Climb' )
bds.raw[case1,"flight_phase"] <- 'takeoff'

case2 <- which(bds.raw$flight_phase=='Landing Roll'|bds.raw$flight_phase=='Taxi'|bds.raw$flight_phase=='Parked')
bds.raw[case2,"flight_phase"] <- 'landing'

case3 <-which(bds.raw$flight_phase=='Approach' | bds.raw$flight_phase=='Descent')
bds.raw[case3,"flight_phase"] = 'inflight'

case4 <- which(bds.raw$flight_phase=='')
bds.raw[case4,"flight_phase"] <- 'unknown'
```
## Change the warned flag as tinyint datatype based on pilot_warned_flag
```{r changingFlags}
warned <- which(bds.raw$pilot_warned_flag=="Y")
notwarned <- which(bds.raw$pilot_warned_flag=='N')

bds.raw[warned,"pilot_warned_flag"] <- 1
bds.raw[notwarned,"pilot_warned_flag"] <- 0
```

## omit flights without flight information
```{r omitEmptyFlightDetails}
bds.raw <- bds.raw[which(bds.raw$airport!='' & bds.raw$model!='' & bds.raw$origin!='' ),]
```


## Question 6 Loading R Dataframe to bird-strikes-DB

### Inserting into conditions table

```{r insertToairportsTable}
conditionsdata <- sqldf::sqldf("SELECT distinct(sky_conditions) as skyCondition from [bds.raw]");
dbWriteTable(birdStrikesDBCon, "conditions", conditionsdata, append = TRUE, row.names = FALSE)
n.conditionsdata <- nrow(conditionsdata)
conditionsdata$cid <- seq(1, n.conditionsdata)
```


### Inserting into airlines table
```{r insertToairlineTable}
airlinesData <- sqldf::sqldf("select distinct(TRIM(airline)) as airlineName, pilot_warned_flag as flag  from [bds.raw]");
dbWriteTable(birdStrikesDBCon, "airlines", airlinesData, append = TRUE, row.names = FALSE)
n.airlinesData <- nrow(airlinesData)
airlinesData$eid <- seq(1, n.airlinesData)
```

### Inserting into airport table

```{r insertToairportTable}
airportsdata <- sqldf::sqldf("select distinct(trim(airport)) as airportName,  origin as state from [bds.raw]")
dbWriteTable(birdStrikesDBCon, "airports", airportsdata, append = TRUE, row.names = FALSE)
n.airportsdata <- nrow(airportsdata)
airportsdata$aid <- seq(1, n.airportsdata)
```

### Inserting into incidents table
```{r insertToIncidentsTable}
incidentsdata <- sqldf::sqldf("select rid, flight_date as depDate, origin, airport, airline, aircraft,flight_phase as flightPhase, altitude_ft as altitude, sky_conditions as conditions, pilot_warned_flag as warned from [bds.raw]");
n.incidentsdata <- nrow(incidentsdata)
for(r in 1:n.incidentsdata)
{
  
  a <- airportsdata$aid[which(airportsdata$airportName == incidentsdata$airport[r])]
  incidentsdata$origin[r] <- a
  
  b <- conditionsdata$cid[which(conditionsdata$skyCondition == incidentsdata$conditions[r])]
  incidentsdata$conditions[r] <- b
  
  c<-airlinesData$eid[which(airlinesData$airlineName == incidentsdata$airline[r])] 
  incidentsdata$airline[r] <- c[[1]][1]
  
}

incidentsdata <- sqldf::sqldf("select rid, depDate , origin, airline, aircraft,flightPhase, altitude, conditions, warned from incidentsdata");

dbWriteTable(birdStrikesDBCon, "incidents", incidentsdata, append = TRUE, row.names = FALSE)

```

## Question 7
```{sql connection=birdStrikesDBCon}
SELECT * from conditions limit 5;
```
```{sql connection=birdStrikesDBCon}
SELECT * from airports limit 5;
```
```{sql connection=birdStrikesDBCon}
SELECT * from airlines limit 5;
```
```{sql connection=birdStrikesDBCon}
SELECT * from incidents limit 5;
```

## Question 8
```{sql connection=birdStrikesDBCon}
select airports.state, COUNT(incidents.rid) as "Number of Incidents" from incidents,airports  where airports.aid= incidents.origin GROUP BY (airports.state) ORDER BY COUNT(incidents.rid) DESC LIMIT 10;
```

## Question 9
```{sql connection=birdStrikesDBCon}
select distinct  airlines.airlineName, COUNT(incidents.rid) as "num_incidents"
from incidents,airlines 
where airlines.eid= incidents.airline
GROUP BY (airlines.airlineName) HAVING num_incidents>
(SELECT AVG(num_incidents) FROM (select distinct  airlines.airlineName, COUNT(incidents.rid) as "num_incidents"
from incidents,airlines 
where airlines.eid= incidents.airline
GROUP BY airlines.airlineName) as average);
```

## Question 10

```{r groupByMonthAndFlightPhase}
no_of_incidents_group_by_month<- dbGetQuery(birdStrikesDBCon,"select distinct MONTH(depDate) as months , COUNT(rid) as num_of_incidents_by_month from incidents group by MONTH(depDate) ORDER BY  MONTH(depDate)")
no_of_incidents_group_by_flightphase <-dbGetQuery(birdStrikesDBCon,"select distinct incidents.flightPhase , COUNT(rid) as num_of_incidents_by_flightphase from incidents group by incidents.flightPhase")
```
```{r}
head(no_of_incidents_group_by_month,6)
head(no_of_incidents_group_by_flightphase,6)
```

## Question 11
```{r plotNoOfIncidentsVsMonths, fig.height = 4}
plot(x = no_of_incidents_group_by_month$months,y = 
       no_of_incidents_group_by_month$num_of_incidents_by_month
,
   xlab = "Month",
   ylab = "No of Bird Strikes",
   main = "Month vs Number of Bird Strike Incidents",
   col =  "blue",
   type = "b")

text(no_of_incidents_group_by_month$months, no_of_incidents_group_by_month$num_of_incidents_by_month, labels=no_of_incidents_group_by_month$num_of_incidents_by_month, cex = 0.5, pos = 4)

legend(x = "topleft", legend=c("Number of bird strikes reported"), col = c("blue"), lty=1:2, cex=0.8, lwd = 2 )


```

## Question 12

```{sql connection=birdStrikesDBCon}
drop procedure if exists add_incident;
```

```{sql connection=birdStrikesDBCon}
create procedure add_incident(
in airport1 text, 
in state1 text, 
in date1 date, 
in airline1 text, 
in aircraft1 text, 
in flightPhase1 text, 
in altitude1 int, 
in condition1 text, 
in warning1 tinyint )

begin

declare cid1 int;
declare aid1 int default 0;
declare rid1 int;
declare eid1 int;

SELECT cid INTO cid1 FROM conditions WHERE conditions.skyCondition = condition1 ;

SELECT aid INTO aid1 FROM airports WHERE airportName = airport1 AND state = state1 ;

SELECT eid INTO eid1 FROM airlines WHERE airlineName= airline1 and flag=warning1;

SELECT (MAX(rid) + 1) INTO rid1 FROM incidents ;

IF aid1 = 0 THEN
SELECT (MAX(aid) + 1) INTO aid1 FROM airports ;
  INSERT INTO airports(aid, airportName, state) VALUES ( aid1, airport1, state1 ) ;
  END IF;

INSERT INTO incidents(rid, depDate,origin, airline, aircraft, flightPhase, altitude, conditions, warned) VALUES ( rid1, date1, aid1, eid1, aircraft1, flightPhase1, altitude1, cid1, warning1 ) ;
end
```



```{sql connection=birdStrikesDBCon}
call add_incident(
'Newyork Airport',
'New York',
'2022-01-01',
'BUSINESS',
'Airplane',
'takeoff',
363,
'Overcast',
0);
```

```{sql connection=birdStrikesDBCon}
select * from incidents where depDate = '2022-01-01'
```

```{sql connection=birdStrikesDBCon}
select * from airports where airportName = 'Newyork Airport'
```
## Disconnecting database
```{r}
dbDisconnect(birdStrikesDBCon)
```
